\phantomsection\addcontentsline{toc}{chapter}{Acknowledgements} % Don't remove me!
\chapter*{\centering{\Large Acknowledgements}}

First and foremost, I would like to extend my deepest gratitude towards Uwe for constantly supporting me throughout the past year. I vividly remember the day I approached you about doing my Honours thesis under your supervision. Your enthusiasm and encouragement convinced me that working with you would be the right choice and it certainly has been so. 

I would also like to extend my thanks to William for being so understanding and generous regarding the completion of this thesis. Your support is also very deeply appreciated.

To all my friends, past and present, your friendship, love and support have made these past four years in university a memorable journey. All of you will remain in my heart and I hope I will remain in yours too.

To Advaitha, Amira, Anindita, Anshu, Anum, Fahim, Gaurav, Jai, Lamiyea, Noshin, Omama, Rohan and Shivang - all of you have been there for me whenever I needed you the most. Words can't describe how much you all mean to me. Life at International House would never have been the same without any of you. Thank you so much. 

To my siblings Shatil, Shaikot and Shawlin, thank you for supporting me through thick and thin and for always coming to my rescue whenever I've needed life advice or emotional support. All three of you have made my path easier with your constant love and guidance. 

Lastly, to my parents. No amount of gratitude can make up for what you have done for me. I am who I am simply because of your upbringing and unwavering love and support. You both mean the world to me. Nothing has made me happier than making the both of you proud. Thank you for everything. 